---
title: "Contrast Coding"
author:
  - name: Filippo Gambarota
date: 2022-01-05
output:
  distill::distill_article:
    self_contained: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Constrast Coding Examples

Difference between `treatment contrast (aka dummy coding)` and `sum contrast`:

* **dummy coding**: one level is coded as reference (0) and each parameter is the difference between the reference and the selected level. For a **two level** factor inside a `lm` we have two parameters the intercept ($\beta_0$) that is the mean when the predictior(s) is 0 so the mean of the reference level and $\beta_1$ that represents the increase of $y$ for a unit increase in $x_i$. A unit increase is the difference between 0 and 1 (two levels) so $\beta_1$ is the mean difference between level 0 and level 1.

Some fake data:

```{r}
y <- c(rnorm(20, 10, 5), rnorm(20, 14, 9)) # simulate a response variable y

sim <- data.frame(y, id = 1:40, gruppo = factor(rep(c("controllo", "sperimentale"), each = 20))) # create a dataset

sim$gruppo_dummy <- sim$gruppo
sim$gruppo_sum_1 <- sim$gruppo
sim$gruppo_sum_2 <- sim$gruppo

# Setting contrasts

contrasts(sim$gruppo_dummy) <- contr.treatment
contrasts(sim$gruppo_sum_1) <- contr.sum
contrasts(sim$gruppo_sum_2) <- c(-0.5, 0.5) # sum contrast with 0.5, -0.5

```

Fitting the models:

```{r}
fit1 <- lm(y ~ gruppo, data = sim)
fit2 <- lm(y ~ gruppo_sum_1, data = sim)
fit3 <- lm(y ~ gruppo_sum_2, data = sim)

coef(fit1)
coef(fit2)
coef(fit3)
```

